---
title: "Materials Project - Batteries"
author: "Maria Pietras"
date: "`r format(Sys.time(), '%d.%M.%Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Executie summary

This report presents an analysis of battery materials from the Materials Project dataset. It explores the components of the most popular batteries, determining that Lithium is by far the most common one. Strong correlations between aspects like Average Voltage and Energy are discovered. Prediction models are build to predict the Gravimetric Capacity of batteries based on other attributes. The first model's performance is unsatisfactory, so another one is build. The second model is based on data that does not contain outliers, and shows a notable improvement. It serves as an experiment to see the significance of the outliers in the dataset.

# Library import

```{r}
library(ggplot2)
library(plotly)
library(dplyr)
library(tidyr)
library(corrplot)
library(tibble)
library(DT)
library(caret)
library(Metrics)
```


# Helper functions

```{r helper}
showTable <- function(table_df, round_digits=2) {
    DT::datatable(table_df, style="bootstrap", filter = "top", rownames = FALSE, extensions = "Buttons", options = list(dom = 'Bfrtip', buttons = c('csv', 'excel', 'pdf', 'print'))) %>% formatRound(names(dplyr::select_if(table_df, is.numeric)), round_digits)
}
```


# Repeatability of results

```{r seed}
set.seed(42)
```

# Data import

MP_Batteries is a dataset prepared in the Materials Project. The Materials Project is a multi-institution, multi-national effort to compute the properties of all inorganic materials and provide the data and associated analysis algorithms for every materials researcher free of charge. The ultimate goal of the initiative is to drastically reduce the time needed to invent new materials by focusing costly and time-consuming experiments on compounds that show the most promise computationally. [source](https://next-gen.materialsproject.org/about)


```{r readData}
data = read.csv('mp_batteries.csv');
showTable(data)
```

The data consists of columns:

  - Battery ID: the identifier of the battery,
  - Battery Formula: the chemical formula of the battery material,
  - Working Ion: the main ion that is responsible for transporting charge in the battery,
  - Charge Formula: the chemical formula of the battery material in the charged state,
  - Formula Discharge: the chemical formula of the battery material in the discharged state,
  - Max Delta Volume: the change in volume in % for a given voltage step using the formula: max(charge, discharge)/min(charge, discharge)-1.
  - Average Voltage: average voltage for a given voltage step,
  - Gravimetric Capacity: the amount of energy per unit mass (mAh/g),
  - Volumetric Capacity: the amount of energy per unit of volume (mAh/cm³),
  - Gravimetric Energy: energy density relative to battery mass (Wh/kg),
  - Volumetric Energy: energy density relative to battery volume (Wh/L),
  - Atomic Fraction Charge: the atomic fraction of components in the charged state,
  - Atomic Fraction Discharge: the atomic fraction of components in the discharged state,
  - Stability Charge: the stability index of the material in the charged state,
  - Stability Discharge: the stability index of the material in the discharged state,
  - Steps: number of distinct voltage steps from full charge to discharge, based on stable intermediate states,
  - Max Voltage Step: maximum absolute difference between adjacent voltage steps.


# Basic statistics of the dataset

## Dimensions

```{r dim}
dim(data)
```

## Summary

```{r summary}
summary(data) %>% knitr::kable()
```

# Missing values

```{r missingData}
any(is.na(data))
```

There are no missing values in the dataset.

# Attribute analysis

## Working Ion

```{r workingIon, echo=FALSE}
ggplot(data, aes(x = Working.Ion)) + geom_bar() + theme_minimal()

```

The most popular working ion is Lithium (Li), with around four times as many occurrences as Calcium (Ca), Magnesium (Mg), Sodium (Na), or Zinc (Zn). Aluminum (Al), Cesium (Cs), Potassium (K), Rubidium (Rb), and Yttrium (Y) belong to minorities.

## Top 10 battery formulas

```{r batteryFormula, echo=FALSE}
top10 <- data %>%
  count(Battery.Formula, sort = TRUE) %>%
  slice_max(n, n = 10) %>%
  pull(Battery.Formula)


top10_data <- data %>%
  filter(Battery.Formula %in% top10)

ggplot(top10_data, aes(x = Battery.Formula)) + geom_bar() +  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

## Density of numerical attributes

```{=html}
<iframe height="500" width="100%" frameborder="no" src="https://mpietr.shinyapps.io/mp_batteries/"> </iframe>
```

Above interactive graphs (hosted on [shinyapps.io](https://mpietr.shinyapps.io/mp_batteries/)) show the density of the numerical attributes. By looking at the shape of the graphs, we can observe that there are outstanding outliers present. 

## Boxplots of numerical attributes

```{r boxplot, fig.width=10, fig.height=20, echo=FALSE}
data %>%
  select_if(is.numeric) %>%
  pivot_longer(cols = everything()) %>%
  ggplot(aes(y = value)) +
  geom_boxplot(fill = "darkblue", alpha = 0.25, color = "black") +
  facet_wrap(~ name, scales = "free", nrow = 3, ncol = 4) +
  labs(x = "Attributes", y = "Values") +
  theme_minimal()
```

The outliers can also be observed on the boxplots. Attributes like Gravimetric Capacity or Volumetric Energy have very high ranges, while attributes like Atomic Friction Charge/Discharge, Max Voltage Step or Stability Charge are confined to smaller ranges (averaging between 0 and 1). Some attributes show small variability with tightly packed plots (like Max Voltage Step or Steps) while others, such as Atomic Friction Discharge, display higher variability.

# Attribute correlation

[Spearman's rank correlation coefficient ($\rho$)](https://en.wikipedia.org/wiki/Spearman%27s_rank_correlation_coefficient) has been chosen for the correlation calculation as it is less sensitive to outliers than, for example, Pearson's. 

The values range from $-1$ to $1$:

  - $\rho=1$: perfect association of rank,
  - $\rho=-1$: perfect negative association of rank,
  - $\rho=0$: no association between ranks.

```{r correlation, echo=FALSE}
M <- data %>%
  select(where(is.numeric)) %>%
  cor(use = "complete.obs", method = "spearman")

df = data.frame(M) %>%
  rownames_to_column() %>%
    pivot_longer(-rowname, names_to="colname")

plot <- ggplot(df, aes(colname, rowname, fill=value)) + 
  geom_tile() +
  scale_fill_gradient2() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  +
  labs(x = "", y = "", fill = "Correlation")

ggplotly(plot)
```

Max Voltage Step and Steps show the highest value of Spearman's Coefficient of around 0.995. Volumetric Energy and Capacity are highly correlated with each other, and a similar situation can be observed for Gravimetric Energy and Capacity. Average Voltage is also significantly correlated with Gravimetric Energy and Volumetric Energy. Atomic Friction Charge and Discharge are correlated with each other as well. Spearman's coefficient is the lowest for Average Voltage and Stability Discharge, amounting to -0.4, wchich means some negative association.
There are also a few pairs that are not correlated at all, for example Max Voltage Step and Stability Charge or Average Voltage and Gravimetric Capacity.

# Trends

The popularization of electrical vehicles (EVs) has increased the demand for fast-charging and high-energy batteries with as high cycle life as possible. The growing environmental awareness in society also contributes to the priority of battery materials research. Scientists explore various new ideas of battery components to improve their performance in different applications.

While traditional batteries like Lithium are still the most popular, with the top 10 battery formulas all containing Lithium, new alternatives are being explored. Using working ions such as Potassium, Magnesium, Aluminum or Sodium could help mitigate ethical, environmental or supply chain risks.

# Prediction

## Attribute selection

Based on the analysis of the attributes and their correlations, I have made the choice to predict Gravimetric Capacity based on Atomic Friction Discharge, Gravimetric Energy and Max Delta Volume. Volumetric Capacity is also highly correlated with Gravimetric Capacity, but since both of the attributes describe capacity (one with relation to mass and the other with relation to volume), it might not be useful to predict one based on the other.

```{r attributeSelection}
model_data <- data %>% select(Atomic.Fraction.Discharge, Gravimetric.Energy, Max.Delta.Volume, Gravimetric.Capacity)
```


## Train and test datasets

The train and test datasets have been split in the ratio of 75% to 25%.

```{r trainTest}
partition <- createDataPartition(model_data$Gravimetric.Capacity, p = 0.75, list = FALSE)

train <- model_data[partition, ]
test <- model_data[-partition, ]
```

## Training

The RandomForest Regression model is repeatedly cross-validated in the process of training.

```{r training}
ctrl <- trainControl(method = "repeatedcv", number = 5, repeats = 5)

model <- train(Gravimetric.Capacity ~ ., data = train,  method = "rf", trControl = ctrl, ntree = 10)

summary(model)
```


## Predictions and scoring

```{r predictions}
predictions <- predict(model, newdata = test)

rmse(test$Gravimetric.Capacity, predictions)
```


```{r predictionsVisualization, echo=FALSE}
plot_data <- data.frame(
  actual = test$Gravimetric.Capacity,
  predicted = predictions
)
 
ggplot(plot_data, aes(x = actual, y = predicted)) +
  geom_point(color = "darkblue") +
  geom_abline(slope = 1, intercept = 0, color = "red") +
  labs(title = "Predicted vs Actual Gravimetric Capacity",
       x = "Actual Gravimetric Capacity (mAh/g)",
       y = "Predicted Gravimetric Capacity (mAh/g)") +
  theme_minimal()
```


The Root Mean Square Error of around above 60 might not be satisfactory as it is around 50% of the median of the dataset.

## Model with removed outliers

To see how much the outliers affect the performance, a new model has been prepared. It was trained on a dataset with outliers removed for Gravimetric Capacity based on [Interquantile Range (IQR)](https://en.wikipedia.org/wiki/Interquartile_range#Outliers).

```{r removedOutliersPrediction}

Q1 <- quantile(data$Gravimetric.Capacity, 0.25)
Q3 <- quantile(data$Gravimetric.Capacity, 0.75)
IQR <- Q3 - Q1
lower_bound <- Q1 - 1.5 * IQR 
upper_bound <- Q3 + 1.5 * IQR

data_no_outliers <- data[data$Gravimetric.Capacity >= lower_bound & 
                         data$Gravimetric.Capacity <= upper_bound, ]
  
new_model_data <- data_no_outliers %>% select(Atomic.Fraction.Discharge, Gravimetric.Energy, Max.Delta.Volume, Gravimetric.Capacity)

partition <- createDataPartition(new_model_data$Gravimetric.Capacity, p = 0.75, list = FALSE)

new_train <- new_model_data[partition, ]
new_test <- new_model_data[-partition, ]

ctrl <- trainControl(method = "repeatedcv", number = 5, repeats = 5)

model <- train(Gravimetric.Capacity ~ ., data = train,  method = "rf", trControl = ctrl, ntree = 10)

summary(model)

new_predictions <- predict(model, newdata = new_test)

rmse(new_test$Gravimetric.Capacity, new_predictions)

```

```{r removedOutliersPredictionsVisualization, echo=FALSE}
plot_data <- data.frame(
  actual = new_test$Gravimetric.Capacity,
  predicted = new_predictions
)

ggplot(plot_data, aes(x = actual, y = predicted)) +
  geom_point(color = "darkblue") +
  geom_abline(slope = 1, intercept = 0, color = "red") +
  labs(title = "Predicted vs Actual Gravimetric Capacity",
       x = "Actual Gravimetric Capacity (mAh/g)",
       y = "Predicted Gravimetric Capacity (mAh/g)") +
  theme_minimal()
```

The removal of outliers improved the RMSE.

## Conclusions

It is possible to predict Gravimetric Capacity based on Atomic Friction Discharge, Gravimetric Energy and Max Delta Volume, but the model performs better without outliers. However, they make a significant part of the dataset. 